---
title: "K-means"
author: "Savitha Murali"
date: today
---

## 1a. K-Means

### Step 1: Load and Prepare the Data (from CSV file)
```{r}
# Use base R to read the CSV file (no tidyverse required)
penguins <- read.csv("palmer_penguins.csv", stringsAsFactors = TRUE)
head(penguins)
```

### Step 1.1: Select Required Columns and Drop Missing Values
```{r}
penguins_filtered <- penguins[, c("bill_length_mm", "flipper_length_mm")]
penguins_filtered <- penguins_filtered[complete.cases(penguins_filtered), ]
cat("Number of rows after dropping NAs:", nrow(penguins_filtered), "\n")
penguins_filtered[sample(nrow(penguins_filtered), 5), ]
```

### Step 1.2: Visualize the Data
```{r}
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)

ggplot(penguins_filtered, aes(x = bill_length_mm, y = flipper_length_mm)) +
  geom_point(color = "steelblue", alpha = 0.7, size = 2) +
  labs(
    title = "Scatter Plot of Bill Length vs Flipper Length",
    x = "Bill Length (mm)",
    y = "Flipper Length (mm)"
  ) +
  theme_minimal()
```

### Step 2: K-Means Clustering from Scratch in R
```{r}
euclidean_dist <- function(a, b) {
  sqrt(rowSums((a - b)^2))
}
```

### Step 2.1: Initialize Centroids
```{r}
set.seed(42)
data_matrix <- as.matrix(penguins_filtered)
K <- 3
initial_indices <- sample(1:nrow(data_matrix), K)
centroids <- data_matrix[initial_indices, ]
print(centroids)
```

### Step 2.2: Run K-Means Algorithm with Image Saving
```{r}
max_iter <- 10
clusters <- rep(0, nrow(data_matrix))

for (iter in 1:max_iter) {
  for (i in 1:nrow(data_matrix)) {
    distances <- apply(centroids, 1, function(centroid) {
      sum((data_matrix[i, ] - centroid)^2)
    })
    clusters[i] <- which.min(distances)
  }

  new_centroids <- centroids
  for (k in 1:K) {
    cluster_points <- data_matrix[clusters == k, ]
    if (nrow(cluster_points) > 0) {
      new_centroids[k, ] <- colMeans(cluster_points)
    }
  }

  cluster_df <- as.data.frame(data_matrix)
  cluster_df$cluster <- as.factor(clusters)
  centroid_df <- as.data.frame(new_centroids)
  colnames(centroid_df) <- c("bill_length_mm", "flipper_length_mm")

  p <- ggplot(cluster_df, aes(x = bill_length_mm, y = flipper_length_mm, color = cluster)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_point(data = centroid_df, aes(x = bill_length_mm, y = flipper_length_mm),
               color = "black", size = 5, shape = 8) +
    scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73")) +
    labs(title = paste("K-Means Iteration", iter),
         x = "Bill Length (mm)",
         y = "Flipper Length (mm)") +
    theme_minimal(base_size = 14) +
    theme(panel.grid.major = element_line(color = "grey90"),
          legend.position = "none")

  print(p)

  ggsave(filename = sprintf("kmeans_iter_%02d.png", iter),
         plot = p + theme(plot.background = element_rect(fill = "white", color = NA)),
         width = 6, height = 5, dpi = 150)

  if (all(new_centroids == centroids)) {
    cat("Converged at iteration", iter, "\n")
    break
  }
  centroids <- new_centroids
}
```

### Step 3: Compare with Built-in KMeans and Evaluate
```{r}
if (!require("cluster")) install.packages("cluster")
if (!require("factoextra")) install.packages("factoextra")
library(cluster)
library(factoextra)

wcss <- numeric()
silhouette_scores <- numeric()

for (k in 2:7) {
  set.seed(42)
  kmeans_model <- kmeans(data_matrix, centers = k, nstart = 10)
  wcss[k] <- sum(kmeans_model$withinss)
  sil <- silhouette(kmeans_model$cluster, dist(data_matrix))
  silhouette_scores[k] <- mean(sil[, 3])
}

wcss_df <- data.frame(K = 2:7, WCSS = wcss[2:7])
sil_df <- data.frame(K = 2:7, Silhouette = silhouette_scores[2:7])
```

### Step 3.2: Plot Elbow Curve
```{r}
ggplot(wcss_df, aes(x = K, y = WCSS)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Elbow Method: WCSS vs K", x = "Number of Clusters (K)", y = "WCSS") +
  theme_minimal()
```

### Step 3.3: Plot Silhouette Scores
```{r}
ggplot(sil_df, aes(x = K, y = Silhouette)) +
  geom_line(color = "darkgreen", size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Silhouette Scores vs K", x = "Number of Clusters (K)", y = "Silhouette") +
  theme_minimal()
```

### Recommendations 

### Elbow Method (WCSS)
The plot showed a sharp drop from K = 2 to K = 3, after which the reduction in WCSS slowed.

This indicates an elbow at K = 3, which often suggests that K = 3 captures the main structure in the data.

### Silhouette Score
The highest average silhouette score was observed at K = 2, indicating that the data points are more clearly and tightly grouped when divided into 2 clusters.

Silhouette scores typically favor compact and well-separated clusters.

### Final Recommendation:
Metric	Suggested K
Elbow Method	3
Silhouette Score	2

Therefore, K = 2 is recommended as the "right" number of clusters, because it yields the best separation and tightest clusters according to the silhouette score, which is a more interpretable metric for cluster quality.

---

## 2a. K-Nearest Neighbors

### Generate Data
```{r}
set.seed(42)
n <- 100
x1 <- runif(n, -3, 3)
x2 <- runif(n, -3, 3)
boundary <- sin(4 * x1) + x1
y <- ifelse(x2 > boundary, 1, 0) |> as.factor()
dat <- data.frame(x1 = x1, x2 = x2, y = y)
```

### Plot with Boundary
```{r}
x1_seq <- seq(-3, 3, length.out = 300)
boundary_curve <- sin(4 * x1_seq) + x1_seq
boundary_df <- data.frame(x1 = x1_seq, x2 = boundary_curve)

ggplot(dat, aes(x = x1, y = x2, color = y)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_line(data = boundary_df, aes(x = x1, y = x2), color = "black") +
  scale_color_manual(values = c("#E69F00", "#56B4E9")) +
  theme_minimal()
```

### Generate Test Set
```{r}
set.seed(99)
n_test <- 100
x1_test <- runif(n_test, -3, 3)
x2_test <- runif(n_test, -3, 3)
boundary_test <- sin(4 * x1_test) + x1_test
y_test <- ifelse(x2_test > boundary_test, 1, 0) |> as.factor()
test_dat <- data.frame(x1 = x1_test, x2 = x2_test, y = y_test)
```

### KNN Function
```{r}
euclidean_distance <- function(a, b) {
  sqrt(sum((a - b)^2))
}

knn_predict <- function(train_x, train_y, test_x, k = 3) {
  predictions <- vector(length = nrow(test_x))
  for (i in 1:nrow(test_x)) {
    distances <- apply(train_x, 1, function(row) euclidean_distance(row, test_x[i, ]))
    nearest_idx <- order(distances)[1:k]
    nearest_labels <- train_y[nearest_idx]
    predictions[i] <- names(sort(table(nearest_labels), decreasing = TRUE))[1]
  }
  return(as.factor(predictions))
}
```

### Evaluate for k = 1 to 30
```{r}
train_x <- dat[, c("x1", "x2")]
train_y <- dat$y
test_x <- test_dat[, c("x1", "x2")]
test_y <- test_dat$y

accuracy_results <- numeric(30)
for (k in 1:30) {
  preds <- knn_predict(train_x, train_y, test_x, k)
  accuracy_results[k] <- mean(preds == test_y) * 100
}

accuracy_df <- data.frame(k = 1:30, accuracy = accuracy_results)
best_k <- which.max(accuracy_results)
best_accuracy <- accuracy_results[best_k]

cat("✅ Optimal k:", best_k, "\n📈 Highest accuracy:", round(best_accuracy, 2), "%\n")
```

### Plot Accuracy vs K
```{r}
ggplot(accuracy_df, aes(x = k, y = accuracy)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(size = 2) +
  geom_vline(xintercept = best_k, linetype = "dashed", color = "red") +
  labs(
    title = "KNN Classification Accuracy vs K",
    x = "Number of Neighbors (k)",
    y = "Accuracy (%)"
  ) +
  theme_minimal(base_size = 14)
```
